---
title: "Merging Code"
output: html_document
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
install.packages("tidytuesdayR")
install.packages("tidyverse")
install.packages("naniar")
require(naniar)
require(ggplot2)
require(stringi)
require(tidyr)
require(dplyr)
require(corrplot)
```

## Getting the data

```{r}
# Getting the Data
transit_cost_2 <- read.csv("transit-1-16-2026.csv")
gdp_data <- read.csv("world-bank-gdp.csv")
country_code <- read.csv("country-code.csv")
urban <- read.csv("urban-pop.csv")
fed <- read.csv("FEDFUNDS.csv")
```

## Cleaning

```{r}
transit2 <- transit_cost_2[1:1007,c("Country", "City", "Start.year", "End.year", "RR.", "Length", "TunnelPer", "Tunnel", "Elevated", "Atgrade", "Stations", "Platform.Length...Meters.", "Year", "Cost.km..2023.dollars.")]

cols <- c("Start.year", "End.year", "RR.", "Length", "TunnelPer", "Tunnel", "Elevated", "Atgrade", "Stations", "Platform.Length...Meters.", "Year",  "Cost.km..2023.dollars.")

for(col in cols){
  transit2[,col] <- as.numeric(transit2[,col])
}

transit2$Years <- transit2$End.year - transit2$Start.year
transit2$real_cost <- transit2$Cost.km..2023.dollars. * transit2$Length
transit2$log_cost <- log(transit2$Cost.km..2023.dollars.)
transit2$log_real_cost <- log(transit2$real_cost)

transit2
```


## Continent

```{r}
transit2 <- transit2 %>%
  mutate(
    Continent = case_when(
      Country %in% c("US", "CA", "MX", "PA", "DR") ~ "North America",
      Country %in% c("BR", "AR", "CL", "CO", "PE", "EC") ~ "South America",
      Country %in% c("GB", "FR", "DE", "ES", "IT", "NL", "NO",
                     "SE", "DK", "FI", "AT", "BE", "BG",
                     "CZ", "GR", "HU", "CH", "RU", "UK", "PL", "PT", "RS", "RO", "UA") ~ "Europe",
      Country %in% c("CN", "JP", "KR", "IN", "HK", "ID",
                     "IL", "IR", "MY", "BD", "TH", "AE",
                     "BH", "KW", "PH", "PK", "QA", "SA", "SG", "TR", "TW", "VN") ~ "Asia",
      Country %in% c("AU", "NZ") ~ "Oceania",
      Country %in% c("EG", "ZA") ~ "Africa",
      TRUE ~ NA_character_
    ),
    Continent = as.factor(Continent)
  )

transit2
```

## GDP and Urban Pop

```{r}
code_map <- c(
  "AR" = "Argentina",
  "AT" = "Austria",
  "AU" = "Australia",
  "BD" = "Bangladesh",
  "BE" = "Belgium",
  "BG" = "Bulgaria",
  "BH" = "Bahrain",
  "BR" = "Brazil",
  "CA" = "Canada",
  "CH" = "Switzerland",
  "CL" = "Chile",
  "CN" = "China",
  "CO" = "Colombia",
  "CZ" = "Czechia",
  "DE" = "Germany",
  "DK" = "Denmark",
  "DR" = "Dominican Republic",
  "EC" = "Ecuador",
  "EG" = "Egypt, Arab Rep.",
  "ES" = "Spain",
  "FI" = "Finland",
  "FR" = "France",
  "GB" = "United Kingdom",
  "GR" = "Greece",
  "HK" = "Hong Kong SAR, China",
  "HU" = "Hungary",
  "ID" = "Indonesia",
  "IL" = "Israel",
  "IN" = "India",
  "IR" = "Iran, Islamic Rep.",
  "IT" = "Italy",
  "JP" = "Japan",
  "KR" = "Korea, Rep.",
  "KW" = "Kuwait",
  "MX" = "Mexico",
  "MY" = "Malaysia",
  "NL" = "Netherlands",
  "NO" = "Norway",
  "NZ" = "New Zealand",
  "PA" = "Panama",
  "PE" = "Peru",
  "PH" = "Philippines",
  "PK" = "Pakistan",
  "PL" = "Poland",
  "PT" = "Portugal",
  "QA" = "Qatar",
  "RO" = "Romania",
  "RS" = "Serbia",
  "RU" = "Russian Federation",
  "SA" = "Saudi Arabia",
  "SE" = "Sweden",
  "SG" = "Singapore",
  "TH" = "Thailand",
  "TR" = "Turkiye",
  "TW" = "Taiwan, China",
  "UA" = "Ukraine",
  "UK" = "United Kingdom",
  "US" = "United States",
  "UZ" = "Uzbekistan",
  "VN" = "Viet Nam",
  "ZA" = "South Africa"
)

get_wb_value <- function(df, transit_code, year) {
  if (is.na(transit_code) || is.na(year)) return(NA)
  
  year_col <- paste0("X", year)
  if (!(year_col %in% names(df))) return(NA)
  
  wb_name <- code_map[transit_code]
  if (is.na(wb_name)) return(NA)
  
  val <- df[df$Country.Name == wb_name, year_col]
  if (length(val) == 0) return(NA)
  
  as.numeric(val[1])
}

transit2$gdp <- mapply(
  function(ctry, yr) get_wb_value(gdp_data, ctry, yr),
  transit2$Country,
  transit2$Year
)

transit2$urban_pop <- mapply(
  function(ctry, yr) get_wb_value(urban, ctry, yr),
  transit2$Country,
  transit2$Year
)

```

## Histograms and missing values plot

```{r histograms}
par(mfrow = c(3, 3))

hist(transit2$real_cost,
     main = "Real Cost",
     xlab = "real_cost")

hist(transit2$log_real_cost,
     main = "Log Real Cost",
     xlab = "log_real_cost")

hist(transit2$Cost.km..2023.dollars.,
     main = "Cost per km",
     xlab = "Cost per km")

hist(transit2$log_cost,
     main = "Log Cost per km",
     xlab = "log_cost")

hist(transit2$Years,
     main = "Years",
     xlab = "Years")

hist(transit2$Stations,
     main = "Stations",
     xlab = "Stations")

hist(transit2$Length,
     main = "Length",
     xlab = "Length")

hist(transit2$Tunnel,
     main = "Tunnel",
     xlab = "Tunnel")

hist(transit2$gdp,
     main = "GDP",
     xlab = "gdp")


#missing values 
gg_miss_var(transit2)

# looking at this plot we decided to ignore tunnelper, platform length, atgrade, and elevated 
#needed to examine is Years was missing randomly

years_missing_summary <- transit2 %>%
  group_by(Continent) %>%
  summarise(
    total_projects = n(),
    missing_years = sum(is.na(Years)),
    prop_missing = missing_years / total_projects
  )
ggplot(years_missing_summary,
       aes(x = reorder(Continent, prop_missing),
           y = prop_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Proportion of Missing Years by Continent",
    x = "Continent",
    y = "Proportion of Projects with Missing Years"
  )
```

## Imputation
```{r imputation}
#imputation
transit2_imp <- transit2 %>%
  group_by(Country) %>%
  mutate(
    End.year = ifelse(is.na(End.year),
                      round(Start.year + mean(Years, na.rm = TRUE)),
                      End.year)
  ) %>%
  ungroup()

transit2_imp$Years <- transit2_imp$End.year - transit2_imp$Start.year

keep <- c("Country", "Start.year", "End.year", "RR.", "Length", "Tunnel", "Stations", "log_cost", "log_real_cost", "Year", "Cost.km..2023.dollars.", "Years", "real_cost", "Continent", "gdp", "urban_pop")

df_clean <- na.omit(transit2_imp[keep])


```

## Overall cost and continents
```{r alison model}
continent_real <- lm(data=df_clean, sqrt(real_cost) ~ Stations + Years + Continent + gdp + Tunnel)
summary(continent_real)
plot(continent_real)
#vif(continent_real)

##removes RR
step_model <- step(continent_real, direction = "both")
summary(step_model)

df_urban_pop <- filter(transit2, is.na(urban_pop))
df_gdp <- filter(transit2, is.na(gdp))
df_cont <- filter(transit2, is.na(Continent))
```

## Testing for spatial and autocorrelation
```{r spatial and auto correlation}

```

# Corr plot---
title: "Merging Code"
output: html_document
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
install.packages("tidytuesdayR")
install.packages("tidyverse")
install.packages("naniar")
require(naniar)
require(ggplot2)
require(stringi)
require(tidyr)
require(dplyr)
require(corrplot)
```

## Getting the data

```{r}
# Getting the Data
transit_cost_2 <- read.csv("transit-1-16-2026.csv")
gdp_data <- read.csv("world-bank-gdp.csv")
country_code <- read.csv("country-code.csv")
urban <- read.csv("urban-pop.csv")
fed <- read.csv("FEDFUNDS.csv")
```

## Cleaning

```{r}
transit2 <- transit_cost_2[1:1007,c("Country", "City", "Start.year", "End.year", "RR.", "Length", "TunnelPer", "Tunnel", "Elevated", "Atgrade", "Stations", "Platform.Length...Meters.", "Year", "Cost.km..2023.dollars.")]

cols <- c("Start.year", "End.year", "RR.", "Length", "Tunnel", "Elevated", "Atgrade", "Stations", "Platform.Length...Meters.", "Year",  "Cost.km..2023.dollars.")

for(col in cols){
  transit2[,col] <- as.numeric(transit2[,col])
}

transit2$Years <- transit2$End.year - transit2$Start.year
transit2$real_cost <- transit2$Cost.km..2023.dollars. * transit2$Length
transit2$log_cost <- log(transit2$Cost.km..2023.dollars.)
transit2$log_real_cost <- log(transit2$real_cost)
transit2$TunnelPer <- as.numeric(gsub("%", "", transit2$TunnelPer))

transit2
```


## Continent

```{r}
transit2 <- transit2 %>%
  mutate(
    Continent = case_when(
      Country %in% c("US", "CA", "MX", "PA", "DR") ~ "North America",
      Country %in% c("BR", "AR", "CL", "CO", "PE", "EC") ~ "South America",
      Country %in% c("GB", "FR", "DE", "ES", "IT", "NL", "NO",
                     "SE", "DK", "FI", "AT", "BE", "BG",
                     "CZ", "GR", "HU", "CH", "RU", "UK", "PL", "PT", "RS", "RO", "UA") ~ "Europe",
      Country %in% c("CN", "JP", "KR", "IN", "HK", "ID",
                     "IL", "IR", "MY", "BD", "TH", "AE",
                     "BH", "KW", "PH", "PK", "QA", "SA", "SG", "TR", "TW", "VN") ~ "Asia",
      Country %in% c("AU", "NZ") ~ "Oceania",
      Country %in% c("EG", "ZA") ~ "Africa",
      TRUE ~ NA_character_
    ),
    Continent = as.factor(Continent)
  )

transit2
```

## GDP and Urban Pop

```{r}
code_map <- c(
  "AR" = "Argentina",
  "AT" = "Austria",
  "AU" = "Australia",
  "BD" = "Bangladesh",
  "BE" = "Belgium",
  "BG" = "Bulgaria",
  "BH" = "Bahrain",
  "BR" = "Brazil",
  "CA" = "Canada",
  "CH" = "Switzerland",
  "CL" = "Chile",
  "CN" = "China",
  "CO" = "Colombia",
  "CZ" = "Czechia",
  "DE" = "Germany",
  "DK" = "Denmark",
  "DR" = "Dominican Republic",
  "EC" = "Ecuador",
  "EG" = "Egypt, Arab Rep.",
  "ES" = "Spain",
  "FI" = "Finland",
  "FR" = "France",
  "GB" = "United Kingdom",
  "GR" = "Greece",
  "HK" = "Hong Kong SAR, China",
  "HU" = "Hungary",
  "ID" = "Indonesia",
  "IL" = "Israel",
  "IN" = "India",
  "IR" = "Iran, Islamic Rep.",
  "IT" = "Italy",
  "JP" = "Japan",
  "KR" = "Korea, Rep.",
  "KW" = "Kuwait",
  "MX" = "Mexico",
  "MY" = "Malaysia",
  "NL" = "Netherlands",
  "NO" = "Norway",
  "NZ" = "New Zealand",
  "PA" = "Panama",
  "PE" = "Peru",
  "PH" = "Philippines",
  "PK" = "Pakistan",
  "PL" = "Poland",
  "PT" = "Portugal",
  "QA" = "Qatar",
  "RO" = "Romania",
  "RS" = "Serbia",
  "RU" = "Russian Federation",
  "SA" = "Saudi Arabia",
  "SE" = "Sweden",
  "SG" = "Singapore",
  "TH" = "Thailand",
  "TR" = "Turkiye",
  "TW" = "Taiwan, China",
  "UA" = "Ukraine",
  "UK" = "United Kingdom",
  "US" = "United States",
  "UZ" = "Uzbekistan",
  "VN" = "Viet Nam",
  "ZA" = "South Africa"
)

get_wb_value <- function(df, transit_code, year) {
  if (is.na(transit_code) || is.na(year)) return(NA)
  
  year_col <- paste0("X", year)
  if (!(year_col %in% names(df))) return(NA)
  
  wb_name <- code_map[transit_code]
  if (is.na(wb_name)) return(NA)
  
  val <- df[df$Country.Name == wb_name, year_col]
  if (length(val) == 0) return(NA)
  
  as.numeric(val[1])
}

transit2$gdp <- mapply(
  function(ctry, yr) get_wb_value(gdp_data, ctry, yr),
  transit2$Country,
  transit2$Year
)

transit2$urban_pop <- mapply(
  function(ctry, yr) get_wb_value(urban, ctry, yr),
  transit2$Country,
  transit2$Year
)

```

## Histograms and missing values plot

```{r histograms}
par(mfrow = c(3, 3))

hist(transit2$real_cost,
     main = "Real Cost",
     xlab = "real_cost")

hist(transit2$log_real_cost,
     main = "Log Real Cost",
     xlab = "log_real_cost")

hist(transit2$Cost.km..2023.dollars.,
     main = "Cost per km",
     xlab = "Cost per km")

hist(transit2$log_cost,
     main = "Log Cost per km",
     xlab = "log_cost")

hist(transit2$Years,
     main = "Years",
     xlab = "Years")

hist(transit2$Stations,
     main = "Stations",
     xlab = "Stations")

hist(transit2$Length,
     main = "Length",
     xlab = "Length")

hist(transit2$Tunnel,
     main = "Tunnel",
     xlab = "Tunnel")

hist(transit2$gdp,
     main = "GDP",
     xlab = "gdp")


#missing values 
gg_miss_var(transit2)

# looking at this plot we decided to ignore platform length, atgrade, and elevated 
#needed to examine is Years was missing randomly

years_missing_summary <- transit2 %>%
  group_by(Continent) %>%
  summarise(
    total_projects = n(),
    missing_years = sum(is.na(Years)),
    prop_missing = missing_years / total_projects
  )
ggplot(years_missing_summary,
       aes(x = reorder(Continent, prop_missing),
           y = prop_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Proportion of Missing Years by Continent",
    x = "Continent",
    y = "Proportion of Projects with Missing Years"
  )
```

## Imputation
```{r imputation}
#imputation
transit2_imp <- transit2 %>%
  group_by(Country) %>%
  mutate(
    End.year = ifelse(is.na(End.year),
                      round(Start.year + mean(Years, na.rm = TRUE)),
                      End.year)
  ) %>%
  ungroup()

transit2_imp$Years <- transit2_imp$End.year - transit2_imp$Start.year

keep <- c("Country", "Start.year", "End.year", "RR.", "Length", "TunnelPer", "Tunnel", "Stations", "log_cost", "log_real_cost", "Year", "Cost.km..2023.dollars.", "Years", "real_cost", "Continent", "gdp", "urban_pop")

df_clean <- na.omit(transit2_imp[keep])


```

## Overall cost and continents
```{r alison model}
continent_real <- lm(data=df_clean, sqrt(real_cost) ~ Stations + RR. + TunnelPer + Years + Continent + gdp + Tunnel)
summary(continent_real)
plot(continent_real)
#vif(continent_real)

##removes RR
step_model <- step(continent_real, direction = "both")
summary(step_model)

df_urban_pop <- filter(transit2, is.na(urban_pop))
df_gdp <- filter(transit2, is.na(gdp))
df_cont <- filter(transit2, is.na(Continent))
```

## Testing for spatial and autocorrelation
```{r spatial and auto correlation}

```

# Corr plot
```{r corr}
numeric_vars <- transit2 %>%
  select(real_cost, Years, Start.year,
         Length, Stations, TunnelPer, urban_pop, gdp)
numeric_vars <- numeric_vars %>%
  filter(complete.cases(.))
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```
```{r corr}
numeric_vars <- transit2 %>%
  select(real_cost, Years, Start.year,
         Length, Stations, TunnelPer, urban_pop, gdp)
numeric_vars <- numeric_vars %>%
  filter(complete.cases(.))
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```